---
title: "An enriched WHO dataset from the DRPRG publication"
author: "Kerri M Malone"
date: "2023-10-19"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}


libs_load <- function(x){
  for( i in x ){
    print(paste0("Checking for library: ", i))
    if(require( i , character.only = TRUE ) ){
      print(paste0(i, " already installed. Loading now"))
    }
    #  require returns TRUE invisibly if it was able to load package
    if( ! require( i , character.only = TRUE ) ){
      print(paste0(i, " not installed. Trying CRAN for install."))
      #  If package was not able to be loaded then re-install
      install.packages( i , dependencies = TRUE )
      require( i , character.only = TRUE )
      paste0(i, " installed and loaded successfully")
    }
    if ( ! require(i, character.only=TRUE) ) {
      paste0(i," could not be installed from CRAN. Trying Bionconductor....")
      BiocManager::install(i)
      require( i , character.only = TRUE )
      paste0(i, " installed and loaded successfully")
    }
    if ( ! require(i, character.only=TRUE) ) {
      paste0(i, "could not be installed. Check manually")
    }
    #  Load package after installing
  }
}

#Load libraries
libs_load(c("data.table", "tidyverse", 
            "ggpubr", "gtable",
            "gtExtras",
            "reshape2",
            "gt"))

`%notin%` <- Negate(`%in%`)
```
This document summarises the data from https://www.microbiologyresearch.org/content/journal/mgen/10.1099/mgen.0.001081. This study was produced by Michael Hall from Zam's group and describes the development of the DRPRG tool that allows for drug resistance prediction using genome graphs.  
  
In doing so, Michael collated and carefully(!) curated an enriched WHO dataset, which is summarised here.



```{r echo = FALSE}
### Data was accessed here [https://doi.org/10.5281/zenodo.7819984] on 20231018.
ROOTDIR = "/Users/kmalone/macbook_m1_backup/github/cryptic-catalogue/drprg-paper-preprint/"

pheno_data = fread(paste0(ROOTDIR, "config/illumina.samplesheet.csv"))


n_samples = length(unique(pheno_data$biosample))

WHO_drugs = pheno_data %>%
  select(-run, -bioproject, -biosample) %>%
  names()

same_biosample_dff_runIDs = pheno_data %>% 
  group_by(biosample) %>%
  mutate(n = n()) %>% 
  filter(n >1) %>% 
  select(biosample) %>% 
  distinct() %>% 
  nrow()

# pheno_data %>%
#   select(-run, -bioproject, -biosample) %>%
#   summarise(across(everything(), ~ list(R = sum(. == "R"), S = sum(. == "S")))) %>%
#   mutate(phenotype = c("R", "S")) %>%
#   gt(rowname_col = "phenotype") %>%
#   gt_theme_nytimes()
```

### Summary:  

* #### There are `r n_samples` unique biosample IDs in the dataset.  
(this excludes `r same_biosample_dff_runIDs` instances of duplicated biosample IDs with different runIDs).  
  
  
* There is at least one drug phenotype for each sample for `r length(WHO_drugs)` drugs: `r WHO_drugs`. 
  
  
A breakdown of R/S phenotypes for the samples can be seen in the plot below:  

```{r warning=FALSE, fig.width = 10, fig.height = 5}
pheno_data %>%
  select(-run, -bioproject, -biosample) %>%
  summarise(across(everything(), ~ list(R = sum(. == "R"), S = sum(. == "S")))) %>%
  mutate(phenotype = c("R", "S")) %>% 
  pivot_longer(-phenotype, names_to = "antibiotic", values_to = "value") %>%
  as.data.frame() %>%
  mutate(value = as.numeric(as.character(value))) %>%
  mutate(antibiotic = case_when(antibiotic == "para-aminosalicylic_acid" ~ "PAS",
                                TRUE ~ antibiotic)) %>%
    ggplot(., aes(x = antibiotic, y = value)) +
    geom_bar(aes(fill = phenotype), position = "dodge",stat = "identity") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0, size = 8)) +
  geom_text(
    stat = "identity", 
    aes(label = value, group = phenotype),
    vjust = -1,
    position = position_dodge(width = 0.9),
    size = 2.5) +
  ylab("# samples") +
  labs(title = "Data used for DRPRG: WHO samples + extra samples",
       subtitle = paste0("n = ", n_samples, " samples"),
       caption = "There is at least one drug phenotype for each sample")

```

The WHO data (minus the "enriched" part) was used to train the WHO catalogue.
The "enriched" dataset minus the WHO data (e.g. the data from other publications that wasn't included in the WHO training data) is as follows:  

```{r echo=FALSE}

who = fread(paste0(ROOTDIR, "config/samplesheets/who.csv"))

not_WHO_data = anti_join(pheno_data, who, by = "run")

not_WHO_data = not_WHO_data %>%
  select(run, biosample, bioproject, everything())

# write.csv(not_WHO_data, file = paste0(ROOTDIR, "validation_set_20231110.csv"),
#                                        quote = F, row.names = F)

n_samples_not_WHO = length(unique(not_WHO_data$biosample))

```


### Summary:  

* #### There are `r n_samples_not_WHO` unique biosample IDs in the dataset.  


* There is at least one drug phenotype for each sample for `r length(WHO_drugs)` drugs: `r WHO_drugs`. 
  
  
A breakdown of R/S phenotypes for the samples can be seen in the plot below:  

```{r warning=FALSE, fig.width = 10, fig.height = 5}
not_WHO_data %>%
  select(-run, -bioproject, -biosample) %>%
  summarise(across(everything(), ~ list(R = sum(. == "R"), S = sum(. == "S")))) %>%
  mutate(phenotype = c("R", "S")) %>% 
  pivot_longer(-phenotype, names_to = "antibiotic", values_to = "value") %>%
  as.data.frame() %>%
  mutate(value = as.numeric(as.character(value))) %>%
  mutate(antibiotic = case_when(antibiotic == "para-aminosalicylic_acid" ~ "PAS",
                                TRUE ~ antibiotic)) %>%
    ggplot(., aes(x = antibiotic, y = value)) +
    geom_bar(aes(fill = phenotype), position = "dodge",stat = "identity") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0, size = 8)) +
  geom_text(
    stat = "identity", 
    aes(label = value, group = phenotype),
    vjust = -1,
    position = position_dodge(width = 0.9),
    size = 2.5) +
  ylab("# samples") +
  labs(title = "Data used for DRPRG: extra samples only",
       subtitle = paste0("n = ", n_samples_not_WHO, " samples"),
       caption = "There is at least one drug phenotype for each sample")

```

How many CRyPTIC samples in the WHO training dataset?  
```{r}
cryptic = fread("/Users/kmalone/macbook_m1_backup/github/cryptic-catalogue/data/CRyPTIC_reuse_table_20231107.csv")

# mismatch_WHO_cryptic_run = cryptic %>%
#   dplyr::rename("run" = ENA_RUN) %>%
#   anti_join(., who, by = "run")
# 
# mismatch_WHO_cryptic_run_dedup = mismatch_WHO_cryptic_run %>%
#   separate(run, into = c("run1", "run2"), sep = "\\.", extra = "merge") %>%
#   pivot_longer(cols = starts_with("run"), names_to = "run_type", values_to = "ENA_RUN") %>%
#   filter(ENA_RUN != "") %>%
#   select(-run_type)
# 
# mismatch_WHO_cryptic_run_dedup %>%
#   dplyr::rename("run" = ENA_RUN) %>%
#   anti_join(., who, by = "run")
# 
# mismatch_WHO_cryptic_ENA = cryptic %>%
#   dplyr::rename("sample" = ENA_SAMPLE) %>%
#   anti_join(., who, by = "sample")


mismatch_pheno_cryptic_run = cryptic %>%
    separate(ENA_RUN, into = c("run1", "run2"), sep = "\\.", extra = "merge") %>%
    pivot_longer(cols = starts_with("run"), names_to = "run_type", values_to = "run") %>%
    filter(run != "") %>%
    anti_join(., pheno_data, by = "run") %>%
  dplyr::rename("biosample" = ENA_SAMPLE)


mismatch_pheno_cryptic_ENA = cryptic %>%
  dplyr::rename("biosample" = ENA_SAMPLE) %>%
  anti_join(., pheno_data, by = "biosample") %>%
  dplyr::rename("run" = ENA_RUN)


mismatch_pheno_cryptic_ENA %>%
    separate(run, into = c("run1", "run2"), sep = "\\.", extra = "merge") %>%
    pivot_longer(cols = starts_with("run"), names_to = "run_type", values_to = "run") %>%
    filter(run != "") %>%
    anti_join(., pheno_data, by = "run") %>% 
  select(run, biosample)


#sites
mismatch_pheno_cryptic_ENA %>%
    separate(run, into = c("run1", "run2"), sep = "\\.", extra = "merge") %>%
    pivot_longer(cols = starts_with("run"), names_to = "run_type", values_to = "run") %>%
    filter(run != "") %>%
    anti_join(., pheno_data, by = "run") %>% 
  dplyr::mutate(x = str_split(UNIQUEID, "\\.subj") %>% map_chr(1)) %>% 
  with(table(.$x))

```

## How was this data created?  


According to the README.md in the data source, the phenotype data can be found in `config/illumina.samplesheet.csv`.  
It was synthesized by summarising the experiments found in `config/samplesheets/` using `workflow/notebook/notepad.ipynb`.
This notebook is large and includes various analyses but the one we are interested in is "Data Collation".

The first steps involved creating a WHO base dataset with two different data sources.  

1. "gentb", which is labelled as WHO correspondence. This is the data underlying Maha Farhat's GenTB paper.

2. "WHO", which I'm assuming is from the latest catalogue. 

3. Lots of data cleaning, resolving discrepancies between the two datasets and extra metadata added using code from https://github.com/mbhall88/WHO-correspondence/blob/main/docs/fill_in_who_samplesheet.py.  


Next, data was sequentially added and cleaned data from various publications:

```{r echo = FALSE}

dataset = c(
           "gentb",
           "trisakil", 
           "smith",
           "peker",
           "merker",
           "finci",
           "leah_bdq",
           "marco_pheno",
           "lempens_acc") 
publication = c(
                "https://genomemedicine.biomedcentral.com/articles/10.1186/s13073-021-00953-4",
                "https://doi.org/10.1080/22221751.2022.2099304",
                "https://pubmed.ncbi.nlm.nih.gov/33055186/",
                "https://doi.org/10.1099/mgen.0.000695",
                "https://doi.org/10.1038/s41467-022-32455-1",
                "https://pubmed.ncbi.nlm.nih.gov/35907429/",
                "https://doi.org/10.1101/2022.12.08.519610",
                "https://doi.org/10.3389/fmicb.2023.1104456",
                "https://doi.org/10.1016/j.ijid.2020.08.042"
                )
pheno_method = c(
                 "various", 
                 "various", 
                 "liquid MGIT 960 system (Bactec MGIT SIRE and PZA package inserts; Becton, Dickinson) and solid 7H10 agar proportion method",
                 "incredibly not noted, only reference in methods is 'All MTB isolates were phenotypically tested for drug susceptibility (phenotypic DST)'",
                 "various",
                 "BACTEC MGIT 960 DST and Sensititre MYCOTB MIC plate (binary results reported)",
                 "BACTEC MGIT 960 DST",
                 "not stated, just that they were 'according to WHO classification'",
                 "LJ slopes and 7H11 plates, proportional")

cbind(dataset,
      publication,
      pheno_method) %>%
  as.data.frame() %>%
  gt() %>%
  gt_theme_nytimes() %>%
  cols_label(
    publication = "reference",
    pheno_method ="phenotype method"
  )
  
```


Then, the following was noted:
"Get the BioProject of all BioSamples with antibiogram data in NCBI. Once I have the BioProject, I can .....download the antibiogram table".  


This resulted in an extra 1073 samples being added to the superset with phenotypes for at least one drug.  

```{r echo = FALSE}
bioproj = c("PRJNA353873",
            "PRJNA413593",
            "PRJNA438921",
            "PRJNA557083",
            "PRJNA650381",
            "PRJNA663350",
            "PRJNA717333",
            "PRJNA824124",
            "PRJNA834625",
            "PRJNA888434")

publication_2 = c(rep("", length(bioproj)))
                  
pheno_method_2 = c("MGIT, MICs listed",
                   "MGIT, MICs listed",
                   "MGIT, MICs listed",
                   "MGIT, MICs listed",
                   "MGIT and proportional agar, MICs listed for both",
                   "MGIT, MICs listed",
                   "96 well plate, MICs listed",
                   "MGIT, MICs listed",
                   "LJ slopes, MICs listed",
                   "MGIT, MICs listed")

cbind(bioproj,
      publication_2,
      pheno_method_2) %>%
  as.data.frame() %>%
  gt() %>%
  gt_theme_nytimes() %>%
  cols_label(
    publication_2 = "reference",
    pheno_method_2 ="phenotype method"
  )

```


All data together:  
```{r echo=FALSE}
df_1 = as.data.frame(cbind(dataset,
      publication,
      pheno_method))

df_2 = as.data.frame(cbind(bioproj,
      publication_2,
      pheno_method_2))

names(df_2)[1:3] = c("dataset", "publication", "pheno_method")

rbind(df_1, df_2)%>%
  as.data.frame() %>%
  gt() %>%
  gt_theme_nytimes() %>%
  cols_label(
    publication = "reference",
    pheno_method ="phenotype method"
  )

```