---
title: "Creating training and validation datasets"
author: "Kerri M Malone"
date: "2023-11-22"
output: html_document
---

```{r}
libs_load <- function(x){
  for( i in x ){
    print(paste0("Checking for library: ", i))
    if(require( i , character.only = TRUE ) ){
      print(paste0(i, " already installed. Loading now"))
    }
    #  require returns TRUE invisibly if it was able to load package
    if( ! require( i , character.only = TRUE ) ){
      print(paste0(i, " not installed. Trying CRAN for install."))
      #  If package was not able to be loaded then re-install
      install.packages( i , dependencies = TRUE )
      require( i , character.only = TRUE )
      paste0(i, " installed and loaded successfully")
    }
    if ( ! require(i, character.only=TRUE) ) {
      paste0(i," could not be installed from CRAN. Trying Bionconductor....")
      BiocManager::install(i)
      require( i , character.only = TRUE )
      paste0(i, " installed and loaded successfully")
    }
    if ( ! require(i, character.only=TRUE) ) {
      paste0(i, "Could not be installed. Check manually with github?")
    }
    #  Load package after installing
  }
}

#Load libraries
libs_load(c("tidyverse", 
            "data.table",
            "readr",
            "UpSetR"))

`%notin%` <- Negate(`%in%`)

```


```{r echo = FALSE}
ROOTDIR = "/Users/kmalone/macbook_m1_backup/github/cryptic-catalogue/"

# Load in files
cryptic = fread(paste0(ROOTDIR, "data/CRyPTIC_reuse_table_20231208.csv"))
mykrobe = fread(paste0(ROOTDIR, "drprg-paper-preprint/mykrobe.20231121.tsv"))
validation = fread(paste0(ROOTDIR, "drprg-paper-preprint/validation_set_20231110.pass.tsv"))

```
We have three datasets at the moment: cryptic, mykrobe and validation.  
Cryptic and mykrobe will be combined to make the training data.  
Will check for any sample dupes between the training data and validation.  


TODO:   
1. Check for any overlap between mykrobe and cryptic biosample IDs before combining. 
2. Phenotype missingness between datasets for now will remain. We can talk about whether to assume susceptibility or not. 
3. Create upset plots to assess phenotype representation between the train and validation data. Can then assess whether to stratify.

1. Make training dataset by combining cryptic and mykrobe:   
```{r}
# First check that biosample IDs are unique to each dataset
mykrobe_ids = mykrobe %>% pull(sample)
cryptic_ids = cryptic %>% pull(ENA_SAMPLE)

training_ids_in_common = intersect(mykrobe_ids, cryptic_ids)
training_ids_in_common_df = as.data.frame(training_ids_in_common)
names(training_ids_in_common_df) = "ena_run"

# write_tsv(training_ids_in_common_df, file = paste0(ROOTDIR, "drprg-paper-preprint/training_data_samples_excluded_duplicated_20231122.tsv"))


```
There are `r length(training_ids_in_common)` biosample IDs in common between mykrobe and cryptic training datasets. These are all site 6/Italian samples that were submitted to cryptic.  

I will remove these from the mykrobe dataset and keep the cryptic entries when merging.  

Now, need to make sure that colnames match between the sets before merging.  

```{r}
drugs = c("amikacin", "bedaquiline", "capreomycin", "ciprofloxacin", "clofazimine", "cycloserine", 
          "delamanid", "ethambutol", "ethionamide", "isoniazid", "kanamycin", "levofloxacin", 
          "linezolid", "moxifloxacin", "ofloxacin", "pas", "pyrazinamide", "rifabutin",
          "rifampicin", "streptomycin")

drug_codes = c("AMI", "BDQ", "CAP","CIP","CFZ", 
               "CS", "DLM", "EMB", "ETH", 
               "INH", "KAN", "LEV", "LZD", 
               "MXF", "OXF", "PAS", "PZA",
               "RIF", "RFB", "STR")

drug_translation = as.data.frame(cbind(drugs, drug_codes))

drug_phenotype_quality = paste0(drugs, "_phenotype_quality")
#write_tsv(drug_translation, file = paste0(ROOTDIR, "drprg-paper-preprint/drug_translation_table_20231122.tsv"))

colnames_order = c("ena_run", "ena_sample", "ena_project", "dataset", drugs, drug_phenotype_quality)


# Find drugs missing in mykrobe and add new column with "U" for undetermined/unknown
mykrobe_drugs = colnames(mykrobe)[4:19]

mykrobe_missing_drugs = setdiff(drug_translation$drugs, mykrobe_drugs)


mykrobe_edited = mykrobe %>%
  dplyr::rename("ena_run" = run) %>%
  dplyr::rename("ena_sample" = sample) %>%
  dplyr::rename("ena_project" = project)

  
for (col_name in mykrobe_missing_drugs) {
    mykrobe_edited[[col_name]] <- "U"
}

# Setting phenotype quality to medium for all mykrobe samples
for (col_name in drug_phenotype_quality) {
    mykrobe_edited[[col_name]] <- "MEDIUM"
}

# Adding dataset column
mykrobe_edited$dataset = "mykrobe"


# Now filter out the 501 overlapping IDs. Will take cryptic results for these as truth.
mykrobe_edited = mykrobe_edited %>%
  filter(ena_sample %notin% training_ids_in_common) %>%
  select(all_of(colnames_order))

```



```{r}
# Now, find drugs missing in cryptic and add new column with "U" for undetermined/unknown
cryptic_edited = cryptic %>% 
  select(ends_with("TYPE"), c("ENA_RUN", "ENA_SAMPLE", ends_with("QUALITY"))) %>%
  rename_all(~ str_remove_all(., "_BINARY_PHENOTYPE")) %>%
  mutate(ena_project = rep("-", nrow(.))) %>%
  dplyr::rename("ena_run" = ENA_RUN) %>%
  dplyr::rename("ena_sample" = ENA_SAMPLE)



cryptic_drug_codes = colnames(cryptic %>% 
  select(ends_with("TYPE")) %>%
  rename_all(~ str_remove_all(., "_BINARY_PHENOTYPE")))

cryptic_drugs = cryptic_drug_codes %>%
  as.data.frame() %>%
  dplyr::rename("drug_codes" = ".") %>%
  merge(., drug_translation, by = "drug_codes") %>%
  pull(drugs)


cryptic_missing_drugs = setdiff(drugs, cryptic_drugs)


names(cryptic_edited)[1:13] = cryptic_drugs


cryptic_edited = cryptic_edited %>%
  rename_all(~ str_remove_all(., "_PHENOTYPE_QUALITY"))

cryptic_drug_phenotype_quality = paste0(cryptic_drugs, "_phenotype_quality")

names(cryptic_edited)[16:28] = cryptic_drug_phenotype_quality

cryptic_missing_phenotype_quality = setdiff(drug_phenotype_quality, cryptic_drug_phenotype_quality)

for (col_name in cryptic_missing_drugs) {
    cryptic_edited[[col_name]] <- "U"
}

# Setting missing phenotype quality scores to U
for (col_name in cryptic_missing_phenotype_quality) {
    cryptic_edited[[col_name]] <- "U"
}

# Adding dataset column
cryptic_edited$dataset = "CRyPTIC_I"

cryptic_edited = cryptic_edited %>%
  select(all_of(colnames_order)) %>%
  mutate_all(~replace_na(., "U"))

```


Merging cryptic and mykrobe:  
```{r}
training_data = as.data.frame(rbind(cryptic_edited, mykrobe_edited))
validation_ids_in_common = intersect(validation$sample, mykrobe$sample)


plot_order = c("rifampicin", "isoniazid", "ethambutol", "pyrazinamide", "amikacin", "kanamycin", "capreomycin", "levofloxacin", "moxifloxacin", "ciprofloxacin", "ofloxacin", "ethionamide", "rifabutin","streptomycin", "pas", "cycloserine", "bedaquiline","clofazimine", "delamanid", "linezolid" )



#looking at the 131 validation samples that are in training:
training_data %>% filter(ena_sample %in% validation_ids_in_common)%>%
    select(-ena_run, -ena_project, -ena_sample) %>%
    reframe(across(everything(), ~ list(R = sum(. == "R"), S = sum(. == "S"), U = sum(. == "U")))) %>%
    mutate(phenotype = c("R", "S", "U")) %>% 
    pivot_longer(-phenotype, names_to = "antibiotic", values_to = "value") %>%
    as.data.frame() %>%
    mutate(value = as.numeric(as.character(value))) %>%
    ggplot(., aes(x = factor(antibiotic, levels = plot_order), y = value)) +
    geom_bar(aes(fill = phenotype), position = "dodge",stat = "identity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0, size = 8)) +
    geom_text(
        stat = "identity", 
        aes(label = value, group = phenotype),
        vjust = -1,
        position = position_dodge(width = 0.9),
        size = 2.5)

# Will remove them from training, won't upset anything to much

training_data = training_data %>%
  filter(ena_sample %notin% validation_ids_in_common)
#write_tsv(training_data, file = paste0(ROOTDIR, "drprg-paper-preprint/training_data_20231122.tsv"))

validation_ids_in_common_df = as.data.frame(validation_ids_in_common)
names(validation_ids_in_common_df) = "ena_run"
# write_tsv(validation_ids_in_common_df, file = paste0(ROOTDIR, "drprg-paper-preprint/training_data_samples_excluded_validation_duplicated_20231122.tsv"))
```


Bar plot for training data. 

```{r}
n_samples = nrow(training_data)

plot_order = c("rifampicin", "isoniazid", "ethambutol", "pyrazinamide", "amikacin", "kanamycin", "capreomycin", "levofloxacin", "moxifloxacin", "ciprofloxacin", "ofloxacin", "ethionamide", "rifabutin","streptomycin", "pas", "cycloserine", "bedaquiline","clofazimine", "delamanid", "linezolid" )


training_data_barplot = training_data %>%
  select(-ena_run, -ena_project, -ena_sample) %>%
  reframe(across(everything(), ~ list(R = sum(. == "R"), S = sum(. == "S"), U = sum(. == "U")))) %>%
  mutate(phenotype = c("R", "S", "U")) %>% 
  pivot_longer(-phenotype, names_to = "antibiotic", values_to = "value") %>%
  as.data.frame() %>%
  mutate(value = as.numeric(as.character(value))) %>%
    ggplot(., aes(x = factor(antibiotic, levels = plot_order), y = value)) +
    geom_bar(aes(fill = phenotype), position = "dodge",stat = "identity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0, size = 8)) +
  geom_text(
    stat = "identity", 
    aes(label = value, group = phenotype),
    vjust = -1,
    position = position_dodge(width = 0.9),
    size = 2.5) +
  ylab("# samples") +
  xlab("") + 
  labs(title = "Training dataset",
       subtitle = paste0("n = ", n_samples, " samples"),
       caption = "There is at least one drug phenotype for each sample") +
  scale_fill_manual(values = c("#ae0e52", "#6d979c", "grey90"))

ggsave(training_data_barplot, filename = paste0(ROOTDIR, "drprg-paper-preprint/training_data_barplot_20231122.pdf"),
       width = 12.2, height = 6.1)

```
Make upset plot for resistant phenotypes by regimen. 
```{r}
# MDR = RIF + INH R
mdr_plotdata = training_data %>%
  filter(rifampicin == "R" & isoniazid == "R")



mdr_columns_to_transform = mdr_plotdata %>%
  select(-ena_project, -ena_run) %>%
  colnames()


mdr_plotdata = mdr_plotdata %>%
  mutate_at(all_of(columns_to_transform), 
            list(~case_when(. == "R" ~ 1,
                             TRUE ~ 0))) %>%
  select(all_of(plot_order))
n_samples_mdr = nrow(mdr_plotdata)

pdf(file=paste0(ROOTDIR, "drprg-paper-preprint/MDR_training_upset_plot_20231122.pdf"), width = 16.15, height = 7.55)
UpSetR::upset(mdr_plotdata, keep.order = T, nsets = 500, nintersects = 84, sets = rev(plot_order), mb.ratio = c(0.5,0.5), order.by = c("freq"), sets.bar.color	
 = "#ae0e52")
grid.text("MDR samples", x = 0.65, y = 0.95,
          gp = gpar(fontsize = 20))
grid.text(paste0("n = ", n_samples_mdr), x = 0.65, y = 0.91,
          gp = gpar(fontsize = 13))
grid.text("interactions with >5 isolates displayed", x = 0.65, y = 0.88,
          gp = gpar(fontsize = 9))
dev.off()


# Pre-XDR TB: caused by an organism that is resistant to isoniazid, rifampin, and a fluoroquinolone OR by an organism that is resistant to isoniazid, rifampin, and a second-line injectable (amikacin, capreomycin, and kanamycin)
# pre-XDR = RIF + INH + FQ R
preXDR_FQ_plotdata = training_data %>%
  filter(rifampicin == "R" & isoniazid == "R" & (moxifloxacin == "R" | ofloxacin == "R" | levofloxacin == "R" | ciprofloxacin == "R"))



preXDR_FQ_columns_to_transform = preXDR_FQ_plotdata %>%
  select(-ena_project, -ena_run) %>%
  colnames()


preXDR_FQ_plotdata = preXDR_FQ_plotdata %>%
  mutate_at(all_of(columns_to_transform), 
            list(~case_when(. == "R" ~ 1,
                             TRUE ~ 0))) %>%
  select(all_of(plot_order))
n_samples_preXDR_FQ = nrow(preXDR_FQ_plotdata)

pdf(file=paste0(ROOTDIR, "drprg-paper-preprint/preXDR_FQ_training_upset_plot_20231122.pdf"), width = 16.15, height = 7.55)
UpSetR::upset(preXDR_FQ_plotdata, keep.order = T, nsets = 500, nintersects = 38, sets = rev(plot_order), mb.ratio = c(0.5,0.5), order.by = c("freq"), sets.bar.color	
 = "#ae0e52")
grid.text("preXDR samples (FQ resistant)", x = 0.65, y = 0.95,
          gp = gpar(fontsize = 20))
grid.text(paste0("n = ", n_samples_preXDR_FQ), x = 0.65, y = 0.91,
          gp = gpar(fontsize = 13))
grid.text("interactions with >5 isolates displayed", x = 0.65, y = 0.88,
          gp = gpar(fontsize = 9))
dev.off()



# pre-XDR = RIF + INH + AG R
preXDR_AG_plotdata = training_data %>%
  filter(rifampicin == "R" & isoniazid == "R" & (kanamycin == "R" | amikacin == "R" | capreomycin == "R"))



preXDR_AG_columns_to_transform = preXDR_AG_plotdata %>%
  select(-ena_project, -ena_run) %>%
  colnames()


preXDR_AG_plotdata = preXDR_AG_plotdata %>%
  mutate_at(all_of(columns_to_transform), 
            list(~case_when(. == "R" ~ 1,
                             TRUE ~ 0))) %>%
  select(all_of(plot_order))

n_samples_preXDR_AG = nrow(preXDR_AG_plotdata)

pdf(file=paste0(ROOTDIR, "drprg-paper-preprint/preXDR_AG_training_upset_plot_20231122.pdf"), width = 16.15, height = 7.55)
UpSetR::upset(preXDR_AG_plotdata, keep.order = T, nsets = 500, nintersects = 36, sets = rev(plot_order), mb.ratio = c(0.5,0.5), order.by = c("freq"), sets.bar.color	= "#ae0e52")
grid.text("preXDR samples (AG resistant)", x = 0.65, y = 0.95,
          gp = gpar(fontsize = 20))
grid.text(paste0("n = ", n_samples_preXDR_AG), x = 0.65, y = 0.91,
          gp = gpar(fontsize = 13))
grid.text("interactions with >5 isolates displayed", x = 0.65, y = 0.88,
          gp = gpar(fontsize = 9))
dev.off()


# Extensively drug-resistant TB (XDR TB) is a rare type of multidrug-resistant tuberculosis (MDR TB) that is resistant to isoniazid, rifampin, a fluoroquinolone, and a second-line injectable (amikacin, capreomycin, and kanamycin) OR isoniazid, rifampin, a fluoroquinolone, and bedaquiline or linezolid.
XDR_plotdata = training_data %>%
  filter(rifampicin == "R" & isoniazid == "R" & (moxifloxacin == "R" | ofloxacin == "R" | levofloxacin == "R" | ciprofloxacin == "R") & (kanamycin == "R" | amikacin == "R" | capreomycin == "R"))



XDR_columns_to_transform = XDR_plotdata %>%
  select(-ena_project, -ena_run) %>%
  colnames()


XDR_plotdata = XDR_plotdata %>%
  mutate_at(all_of(columns_to_transform), 
            list(~case_when(. == "R" ~ 1,
                             TRUE ~ 0))) %>%
  select(all_of(plot_order))

n_samples_XDR= nrow(XDR_plotdata)

pdf(file=paste0(ROOTDIR, "drprg-paper-preprint/XDR_FQ_AG_training_upset_plot_20231122.pdf"), width = 16.15, height = 7.55)
UpSetR::upset(XDR_plotdata, keep.order = T, nsets = 500, nintersects = 39, sets = rev(plot_order), mb.ratio = c(0.5,0.5), order.by = c("freq"), sets.bar.color	
 = "#ae0e52")
grid.text("XDR samples (FQ + AG resistant)", x = 0.65, y = 0.95,
          gp = gpar(fontsize = 20))
grid.text(paste0("n = ", n_samples_XDR), x = 0.65, y = 0.91,
          gp = gpar(fontsize = 13))
grid.text("interactions with >3 isolates displayed", x = 0.65, y = 0.88,
          gp = gpar(fontsize = 9))
dev.off()





XDR_BL_plotdata = training_data %>%
  filter(rifampicin == "R" & isoniazid == "R" & (moxifloxacin == "R" | ofloxacin == "R" | levofloxacin == "R" | ciprofloxacin == "R") & (bedaquiline == "R" | linezolid == "R"))



XDR_BL_columns_to_transform = XDR_BL_plotdata %>%
  select(-ena_project, -ena_run) %>%
  colnames()


XDR_BL_plotdata = XDR_BL_plotdata %>%
  mutate_at(all_of(columns_to_transform), 
            list(~case_when(. == "R" ~ 1,
                             TRUE ~ 0))) %>%
  select(all_of(plot_order))

n_samples_XDR_BL = nrow(XDR_BL_plotdata)

pdf(file=paste0(ROOTDIR, "drprg-paper-preprint/XDR_FQ_BDQ_LZD_training_upset_plot_20231122.pdf"), width = 16.15, height = 7.55)
UpSetR::upset(XDR_BL_plotdata, keep.order = T, nsets = 500, nintersects = 31, sets = rev(plot_order), mb.ratio = c(0.5,0.5), order.by = c("freq"), sets.bar.color	
 = "#ae0e52")
grid.text("XDR samples (FQ + BDQ/LZD resistant)", x = 0.65, y = 0.95,
          gp = gpar(fontsize = 20))
grid.text(paste0("n = ", n_samples_XDR_BL), x = 0.65, y = 0.91,
          gp = gpar(fontsize = 13))
grid.text("interactions with >2 isolates displayed", x = 0.65, y = 0.88,
          gp = gpar(fontsize = 9))
dev.off()
```
Summary text file. 
```{r}
file_conn <- file(paste0(ROOTDIR, "drprg-paper-preprint/summary_training_validation_datasets_20231122.txt"), open = "w")

content <- c("Training data: 25033 samples, file = training_data_20231122.tsv",
             "Validation data: 8914 samples, file = validation_set_20231110.pass.tsv",
             "Training data generated from CRyPTIC data (CRyPTIC_reuse_table_20231107.csv) and mykrobe data (mykrobe.20231121.tsv). 501 samples removed due to duplication between the sets.",
             "131 samples removed from the resulting training data as they are part of the validation set")
writeLines(content, file_conn)
close(file_conn)
```